Red Wine Quality Analysis by Jorge Riera
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(ggthemes)
```

```{r echo=FALSE, Load_the_Data}
wine <- read.csv('wineQualityReds.csv')
```

The following study explores various characteristics of red wine including pH, 
citric acid, and residual sugars to determine which factors could be used to
predict quality. These characteristics are compared to the ratings of three 
wine experts. The relationships between these variables are explored throughout 
this study.

# Univariate Plots Section

```{r echo=FALSE, Data_Summary}
dim(wine)

str(wine)

summary(wine)
```

This dataset consists of 13 variables with 1599 observations.

```{r}
#Create function to faciliate plotting of histograms in ggplot
plot.hist <- function(x.var, df, bw){
  ggplot(df, aes_string(x = x.var)) +
    geom_histogram(binwidth = bw)
}

#Create functions to determine which values are outliers for given
#variable
upper.outlier.calc <- function(x.var, df){
  with(df, quantile(x.var, 0.75) + (1.5 * (quantile(x.var, 0.75) - 
                                             quantile(x.var, 0.25))))}

lower.outlier.calc <- function(x.var, df){
  with(df, quantile(x.var, 0.25) - (1.5 * (quantile(x.var, 0.75) - 
                                             quantile(x.var, 0.25))))}
```

```{r echo=F, message=F, warning=F, Univariate_Plots_fixed.acidity}
p1 <- plot.hist('fixed.acidity', wine, 0.4) + 
  scale_x_continuous(breaks = seq(0, 16, 1)) +
  labs(y = 'Fixed Acidity (g / dm^3)', x = 'Count',
          title = "Fixed Acidity distribution") 

p2 <- ggplot(wine, aes(y = fixed.acidity, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = fixed.acidity), alpha = 0.3) +
  labs(y = 'Fixed Acidity (g / dm^3)', x = '',
          title = "Fixed Acidity distribution") +
  coord_flip()

grid.arrange(p1, p2)

upper.outlier.calc(wine$fixed.acidity, wine)

with(wine, summary(fixed.acidity))
```

Most wine have a fixed acidity between 6 and 11. There is a slight skew to the 
right with most wines having a fixed acidity of 11 or lower. There are multiple outliers in this distribution where the wines have fixed acidities higher than 
12.35.

```{r echo=F, message=F, warning=F, Univariate_Plots_volatile.acidity}
p1 <- plot.hist('volatile.acidity', wine, 0.01) 

p2 <- ggplot(wine, aes(y = volatile.acidity, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = volatile.acidity), alpha = 0.3) +
  labs(y = 'Volatile Acidity (g/L)', x = '',
          title = "Volatile acidity distribution") +
  coord_flip()

grid.arrange(p1, p2)

upper.outlier.calc(wine$volatile.acidity, wine)

with(wine, summary(volatile.acidity))
```

The wine with the lowest volatile acidity has a score of 0.12 and the highest 
has 1.58. Above, I plot the main body of volatile acidity, trimming those with 
the highest levels. There appears to be a bi modal distribution. There are 
multiple outliers in this distribution where the wines have volatile acidities 
higher than 1.015.

```{r echo=F, message=F, warning=F, Univariate_Plots_citric.acid}
plot.hist('citric.acid', wine, 0.03)

plot.hist('citric.acid', wine, 0.03) + 
  scale_x_log10(limits = c(0.05, 1))

ggplot(wine, aes(y = citric.acid, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = citric.acid), alpha = 0.3) +
  labs(y = 'Citric Acid (g / dm^3)', x = '',
          title = "Citric Acid distribution") +
  coord_flip()

upper.outlier.calc(wine$citric.acid, wine)

with(wine, summary(citric.acid))

table(wine$citric.acid)
```

I transformed the long tail data to better understand the distribution of citric 
acidity. The wine with the lowest citric acid has a score of 0 while the highest
has a score of 1. Most values lie between 0.09 and 0.42. The wine with the 
citric acid level of 1 is the only outlier in this distribution.

```{r echo=F, message=F, warning=F, Univariate_Plots_residual.sugar}
plot.hist('residual.sugar', wine, 0.1)

plot.hist('residual.sugar', wine, 0.1) +
  coord_cartesian(xlim = c(0.5, 3.5))

ggplot(wine, aes(y = residual.sugar, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = residual.sugar), alpha = 0.3) +
  labs(y = 'Residual Sugar (g / dm^3)', x = '',
          title = "Residual Sugar distribution") +
  coord_flip()

upper.outlier.calc(wine$residual.sugar, wine)

with(wine, summary(residual.sugar))

```

Most wines have a residual sugar level value between 1.9 and 2.6. I plotted 
those values that lie in this range and they appear to be normally distributed. 
The wine with the lowest residual sugar has a value of 0.9 and the highest has a 
score of 15.5. There are multiple outliers in this distribution where the wines 
have residual sugar levels higher than 3.65.

```{r echo=F, message=F, warning=F, Univariate_Plots_chlorides}
plot.hist('chlorides', wine, .001)

plot.hist('chlorides', wine, .001) +
  coord_cartesian(xlim = c(0.01, 0.15))

ggplot(wine, aes(y = chlorides, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = chlorides), alpha = 0.3) +
  labs(y = 'Chlorides (g / dm^3)', x = '',
          title = "Chlorides distribution") +
  coord_flip()

upper.outlier.calc(wine$chlorides, wine)

lower.outlier.calc(wine$chlorides, wine)

with(wine, summary(chlorides))
```

The chloride levels for the wines in the dataset range from 0.012 to 0.611. Most 
values lie between 0.07 and 0.09. I plotted the chloride levels for wines that 
had values within this range. These values appear to have a normal distribution. 
There are multiple outliers in this distribution where the wines have chloride 
levels higher than 0.12. There are also a handful of wines in the lower end of 
the distribution that are outliers with chloride levels less than 0.04.

```{r echo=F, message=F, warning=F, Univariate_Plots_free.sulfur.dioxide}
plot.hist('free.sulfur.dioxide', wine, 1)

plot.hist('free.sulfur.dioxide', wine, .05) + scale_x_log10()

ggplot(wine, aes(y = free.sulfur.dioxide, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = free.sulfur.dioxide), alpha = 0.3) +
  labs(y = 'Free Sulfur Dioxide (mg / dm^3)', x = '',
          title = "Free Sulfur Dioxide distribution") +
  coord_flip()

upper.outlier.calc(wine$free.sulfur.dioxide, wine)

with(wine, summary(free.sulfur.dioxide))

with(wine, table(free.sulfur.dioxide))
```

Transforming the plot of free sulfur dioxide reveals a bi modal distribution. 
Most wines have values between 7 and 21. The lowest value is 1 and the highest 
is 72. There are multiple outliers in this distribution where the wines have 
free sulfur dioxide levels higher than 42.

```{r echo=F, message=F, warning=F, Univariate_Plots_total.sulfur.dioxide}
plot.hist('total.sulfur.dioxide', wine, 5)

plot.hist('total.sulfur.dioxide', wine, .02) + scale_x_log10()

ggplot(wine, aes(y = total.sulfur.dioxide, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = total.sulfur.dioxide), alpha = 0.3) +
  labs(y = 'Total Sulfur Dioxide (mg / dm^3)', x = '',
          title = "Total Sulfur Dioxide distribution") +
  coord_flip()

upper.outlier.calc(wine$total.sulfur.dioxide, wine)

with(wine, summary(total.sulfur.dioxide))

with(wine, table(total.sulfur.dioxide))
```

Total sulfur dioxide levels range between 6 and 289. Most values lie between 22 
and 62. The maximum value is 289 and the minimum is 6. There are multiple 
outliers in this distribution where the wines have total sulfur dioxide levels 
higher than 122.

```{r echo=F, message=F, warning=F, Univariate_Plots_density}
p1 <- plot.hist('density', wine, 0.0005) +
  labs(y = 'Density (g / cm^3)', x = 'Count',
          title = "Density distribution") 

p2 <- ggplot(wine, aes(y = density, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = density), alpha = 0.3) +
  labs(y = 'Density (g / cm^3)', x = '',
          title = "Density distribution") +
  coord_flip()

upper.outlier.calc(wine$density, wine)

lower.outlier.calc(wine$density, wine)

grid.arrange(p1, p2)

with(wine, summary(density))
```

The density of the wine's in the dataset range from 0.9901 to 1.004. Most wines 
lie between 0.9956 and 0.9978. There are multiple outliers in this distribution 
where the wines have densities that are either higher than 1.001187 or lower 
than 0.9922475.

```{r echo=F, message=F, warning=F, Univariate_Plots_pH}
plot.hist('pH', wine, .01) + scale_x_continuous(limits = c(3, 4)) 

ggplot(wine, aes(y = pH, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = pH), alpha = 0.3) +
  labs(y = 'pH', x = '',
          title = "pH distribution") +
  coord_flip()

upper.outlier.calc(wine$pH, wine)

lower.outlier.calc(wine$pH, wine)


with(wine, summary(pH))

with(wine, table(pH))
```

The pH levels of wine range from 2.74 to 4.01. Most have values between 3.21 and 
3.4. There are multiple outliers in this distribution where the wines have pH 
levels that are either higher than 3.685 or lower than 2.925.

```{r echo=F, message=F, warning=F, Univariate_Plots_sulphates}
plot.hist('sulphates', wine, .01)

plot.hist('sulphates', wine, .01) +
  scale_x_continuous(limits = c(0.25, 1.5))

ggplot(wine, aes(y = sulphates, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = sulphates), alpha = 0.3) +
  labs(y = 'Sulphates (g / dm3)', x = '',
          title = "Sulphates distribution") +
  coord_flip()

upper.outlier.calc(wine$sulphates, wine)

with(wine, summary(sulphates))

with(wine, table(sulphates))
```

Sulphate levels range from 0.33 to 2.00. Most values lie between 0.55 and 0.73. 
There is a right skew in the distribution of sulphate levels. There are multiple outliers in this distribution where the wines have sulphate levels that are 
higher than 1.

```{r echo=F, message=F, warning=F, Univariate_Plots_alcohol}
p1 <- plot.hist('alcohol', wine, .25) +
  labs(y = 'Alcohol (% by volume)', x = 'Count',
          title = "Alcohol distribution") 

p2 <- ggplot(wine, aes(y = alcohol, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = alcohol), alpha = 0.3) +
  labs(y = 'Alcohol (% by volume)', x = '',
          title = "Alcohol distribution") +
  coord_flip()

upper.outlier.calc(wine$alcohol, wine)

grid.arrange(p1, p2)

with(wine, summary(alcohol))
```

Most wines have alcohol levels between 9.5 and 11.1. The minimum is 8.4 and the 
maximum value is 14.9. There is a right skew in the distribution of alcohol 
levels. There are multiple outliers in this distribution where the wines have 
alcohol levels that are higher than 13.5.

```{r echo=F, message=F, warning=F, Univariate_Plots_quality}
ggplot(data = wine, aes(x = factor(quality))) + geom_bar()

with(wine, summary(quality))

with(wine, table(quality))
```

The quality ratings of wine range from 3 to 8. Most wines having a rating of 5 
or 6. 

# Univariate Analysis

### Structure of dataset

The dataset contains 1599 types of wines each with 11 independent variables that 
are based on physicochemical tests and one dependent variable. The dependent 
variable is the measurement of quality which is based on sensory data provided 
by wine experts.

### Main features of interest

The univariate analysis of the red wine dataset revealed interesting 
distributions in alcohol, fixed acidity, and sulphates. Each of these variables 
has long tails which I will investigate further in my bi-variate analysis. I 
suspect that these variables have an influence on the quality of the wine.

### Other features of interest

The other two variables that I will examine in more detail are citric acid and 
total sulfur dioxide. Citric acid is added to finished wines to increase acidity 
and give the wine a 'fresh flavor.' If too much is added, it can increase the 
formation of volatile acid which gives wines a vinegar taste. There may be a 
'sweet spot' for this variable which I will explore in more detail. Plotting the 
values of total sulfur dioxide revealed a bi modal distribution which may be 
related to the quality of the wine. 

###Grouping Wines Based on Quality

```{r echo=F, message=F, warning=F}
#Label wines as good or bad based on quality rating
cond <- wine$quality >= 6
wine$binary.quality <- ifelse(cond, 'Good', 'Bad')

```

```{r echo=F, message=F, warning=F}
ggplot(data = wine, aes(x = binary.quality)) + geom_histogram(stat="count") + 
  xlab('Quality') + ylab('Count') + ggtitle("Quality of Red Wine's in Dataset")

ggplot(data = wine, aes(x = binary.quality, y = volatile.acidity)) + 
  geom_boxplot()
```

I added an additional variable called 'binary.quality' that will divide the 
wines into two categories- good or bad. Wines that have a quality rating of 6 or 
higher are deemed good. Any wine with a rating less than 6 is considered bad. 
Creating these two groups will help to uncover patterns from a bi-variate and 
multivariate analysis. 

# Bivariate Plots Section

### Pair Plot

```{r echo=F, message=F, warning=F, Bivariate_Plots_pair.plot}
ggpairs(wine, columns = 2:13, 
        upper=list(continuous = wrap("cor", size=3), axisLabels='show'))
```

A pair-plot of the variables in the dataset reveals that alcohol, volatile 
acidity, and citric acid seem to have a strong relationship with the quality of 
a wine. I will explore these relationships further. 

```{r echo=F, message=F, warning=F, Bivariate_Plots_fixed.acidity}
ggplot(data = wine, aes(x = binary.quality, y = volatile.acidity, 
                        fill = binary.quality)) + geom_boxplot(alpha = 1/5)  + scale_fill_brewer(palette="Set1") + 
  ggtitle('Volatile Acidity by Wine Quality') + xlab("Wine Quality") + 
  ylab("Volatile Acidity") + labs(fill = 'Quality')  + 
  theme(plot.title = element_text(hjust = 0.5))

by(wine$volatile.acidity, wine$binary.quality, summary)

ggplot(aes(x = volatile.acidity, y = (..count../sum(..count..) * 100)),
       data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks  = seq(0, 1.6, .1)) + 
  coord_cartesian(xlim = c(0, 1.6)) +  
  ggtitle('Volatile Acidity by Wine Quality') +
  xlab("Volatile Acidity") +
  ylab("% of wines with given volatile acidity") + labs(color="Quality") +
  theme(plot.title = element_text(hjust = 0.5))
```

The boxplot above shows that the variance for volatile acidity in good and bad 
wines is similar. Bad wines have a higher mean and median. The mean and 
median volatile acidity for good wines was 0.4741 and 0.46 respectively. The 
mean and median volatile acidity for bad wines was 0.5895 and 0.59 respectively. 
A frequency polygon of volatile acidity by wine quality shows that the higher 
the volatile acidity, the more likely a wine is to be of bad quality. Most wines 
above a volatile acidity of 0.6 were of bad quality.

```{r echo=F, message=F, warning=F, Bivariate_Plots_sulphates}
ggplot(data = wine, aes(x = binary.quality, y = sulphates, 
                        fill = binary.quality)) + geom_boxplot(alpha = 1/5)  +
    scale_fill_brewer(palette="Set1") + 
  ggtitle('Sulphate level by Wine Type') + 
  xlab("Wine Quality") + ylab("Sulphates") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(wine$sulphates, wine$binary.quality, summary)

ggplot(aes(x = sulphates, y = (..count../sum(..count..) * 100)), data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks  = seq(0.33,2,0.1)) + 
  ggtitle('Sulphate Level by Wine Quality') + 
  coord_cartesian(xlim = c(0.33,2)) +
  xlab("Sulphate Level") +
  ylab("% of wines with given Sulphate Level") + labs(color="Quality") + 
  theme(plot.title = element_text(hjust = 0.5))
```

Sulphate levels also vary between wines of different qualities. Good quality 
wines had higher mean and median sulphate levels. The mean and median values for 
good wines were 0.6926 and 0.66 respectively. Wines of poor quality had mean and 
median sulphate levels of 0.6185 and 0.58 respectively. A frequency polygon of 
sulphate levels by wine quality reveals that those with higher levels have a 
greater likelihood of being good wines. Wines that had sulphate levels of 0.63 
or higher tend to be of good quality. 

```{r echo=F, message=F, warning=F, Bivariate_Plots_alcohol}
ggplot(data = wine, aes(x = binary.quality, y = alcohol, 
                        fill = binary.quality)) + geom_boxplot(alpha = 1/5)  + scale_y_continuous(breaks = seq(8, 14, 1)) + coord_cartesian(ylim = c(8,14)) + 
    scale_fill_brewer(palette="Set1") + 
  ggtitle('Alcohol Content by Wine Type') + xlab("Wine Quality") + 
  ylab("Alcohol Content") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(wine$alcohol, wine$binary.quality, summary)

ggplot(aes(x = alcohol, y = (..count../sum(..count..) * 100)), data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks =seq(8,15, 0.5)) + 
  ggtitle('Alcohol content by Wine Quality') + xlab("Alcohol") +
  ylab("% of wines with given alcohol content") + 
  labs(color="Quality") + theme(plot.title = element_text(hjust = 0.5))
```

Alcohol content varies depending on the quality of the wine. Wines of good 
quality have a higher variance as well as a higher mean and median. The mean and 
median alcohol content for good wines was 10.86 and 10.8 respectively. Bad wines 
had a mean alcohol content of 9.926 and a median of 9.7. A frequency polygon of 
alcohol content by wine quality shows that wines with alcohol levels above 10.25 
have a higher likelihood of being of good quality. 

```{r echo=F, message=F, warning=F, Bivariate_Plots_total.sulfur.dioxide}
ggplot(data = wine, aes(x = binary.quality, y = total.sulfur.dioxide, 
                        fill = binary.quality)) + geom_boxplot(alpha = 1/5)  + scale_y_continuous(breaks = seq(0, 180, 40)) + 
  coord_cartesian(ylim = c(6, 180)) + scale_fill_brewer(palette="Set1") + 
  ggtitle('Total Sulfur Dioxide by Wine Type') + xlab("Wine Quality") +
  ylab("Total Sulfur Dioxide") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(wine$total.sulfur.dioxide, wine$binary.quality, summary)

ggplot(aes(x = total.sulfur.dioxide, y = (..count../sum(..count..) * 100)), 
       data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks =seq(0, 160, 20)) + 
  coord_cartesian(xlim = c(0, 160)) + 
  ggtitle('Total Sulfur Dioxide Level by Wine Quality') +
  xlab("Total Sulfur Dioxide") +
  ylab("% of wines with given Total Sulfur Dioxide") + 
  labs(color="Quality") + theme(plot.title = element_text(hjust = 0.5))

```

Wines of poor quality have a larger variance in total sulfur dioxide levels as 
well as a higher mean and median. Wines of bad quality have a median of 45 and a 
mean of 54.65. Wines of good quality have a mean of 39.35 and a median of 33. A 
plot of the frequency of total sulfur dioxide levels by wine quality suggests 
that wines with levels above 80 are more likely to be of bad quality. 

```{r echo=F, message=F, warning=F, Bivariate_Plots_citric.acid}
ggplot(data = wine, aes(x = binary.quality, y = citric.acid, 
                        fill = binary.quality)) + geom_boxplot(alpha = 1/5)  + scale_y_continuous(breaks = seq(0, 1, .1)) + coord_cartesian(ylim = c(0, 1)) + 
    scale_fill_brewer(palette="Set1") + 
  ggtitle('Citric Acid by Wine Quality') + xlab("Quality") + 
  ylab("Citric Acid") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(wine$citric.acid, wine$binary.quality, summary)

ggplot(aes(x = citric.acid, y = (..count../sum(..count..) * 100)), 
       data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks = seq(0, 1, .1)) + 
  coord_cartesian(xlim = c(0, 1)) + ggtitle('Citric Acid by Wine Quality') + xlab("Citric Acid") + ylab("% of Wines with given Citric Acid Level") + labs(color="Quality") + theme(plot.title = element_text(hjust = 0.5)) 
```

A box plot of the citric acid content of wines by quality reveals that good 
wines have a higher median and variance than bad wines. Good wines also have a 
higher mean citric acid. The mean and median levels for good wines are 0.46 and 
0.31 respectively. Bad wines have a mean of 0.2378 and a median of 0.22. 
Plotting the frequency of citric acid levels by wine quality suggests that good 
wines have higher citric acid levels. One notable aspect of this data is that it 
does not show that at high levels, citric acid can affect the quality of the 
wine. Citric acid can increase the formation of volatile acid. One would expect 
that there would be a resurgence of bad wines at higher levels. 

# Bivariate Analysis

### Relationships between features

```{r echo=F, message=F, warning=F}
ggplot(data = wine, aes(x = volatile.acidity, y = alcohol)) +
  geom_point(alpha = 1/4, color = '#F79420') + 
  ggtitle('Alcohol vs. Fixed Acidity') + xlab("Fixed Acidity") + ylab("Alcohol")

cor(wine$alcohol, wine$fixed.acidity)
```

A bi-variate analysis of the dataset reveals that good wines have tend to have a 
higher alcohol content as well as lower levels of volatile acidity. These two 
variables do not seem to have a significant correlation with each other, but I 
will explore this relationship in more detail in my multivariate analysis. 

### Other Observations

One interesting relationship that wasn't a main feature of interest, is that of 
fixed acidity and density. These two variables seem to have a significant 
correlation. This will be explored further. 

### Strongest Relationship

```{r echo=F, message=F, warning=F}
cor(wine$pH, wine$fixed.acidity)
```

Fixed acidity and pH have the strongest correlation among all the variables 
analyzed in this dataset. The correlation coefficient for these two variables is 
-0.6829782. This would make sense because the pH scale is a measure of how basic 
or acidic a substance is. One would expect a high correlation for these two 
variables. 

# Multivariate Plots Section

```{r echo=F, message=F, warning=F, Multivariate_Plots_fixed.acidity.v.density}
ggplot(data = wine, aes(x = fixed.acidity, y = density)) +
  geom_point(aes(color = binary.quality)) + 
  scale_x_continuous(breaks = seq(4, 16, 1)) + 
  scale_y_continuous(breaks = seq(0.99, 1.1, .002)) +
ggtitle('Fixed Acidity vs. Density') + xlab("Fixed Acidity") + ylab("Density") +
  theme(plot.title = element_text(hjust = 0.5))

cor(wine$fixed.acidity, wine$density)
```

In the previous section, I noted that density and fixed acidity seem to have a 
significant correlation. I examined this relationship further by creating a 
scatter plot of the two variables and labeling each point based on the quality 
of the underlying wine. It seems like at lower fixed acidity levels, good wines 
tend to have lower densities.  

```{r echo=F, message=F, warning=F, Multivariate_Plots}
ggplot(data = wine, aes(x = alcohol, y = volatile.acidity)) + 
  geom_point(aes(color = binary.quality))
```

It seems like the higher the alcohol content and lower the volatile acidity of a 
wine, the more likely it is to be of good quality. 

```{r echo=F, message=F, warning=F}
ggplot(data = wine, aes(x = alcohol, y = sulphates)) + 
  geom_point(aes(color = binary.quality))
```

Higher sulphate levels and alcohol content also seem to be indicators of a good 
wine.

```{r echo=F, message=F, warning=F}
ggplot(data = wine, aes(x = alcohol, y = citric.acid)) + 
  geom_point(aes(color = binary.quality))
```

Similarly, wines with higher alcohol and citric acid levels tend to be of good 
quality.

# Multivariate Analysis

Wines with higher alcohol, volatile acidity, and sulphate levels tend to be of 
better quality. One could use these features to construct a linear model to 
predict the quality of a red wine. 

# Building a Model Using the Red Wine Dataset

```{r echo=F, message=F, warning=F}
m1 <- lm(I(quality ~ alcohol), data = wine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)

mtable(m1, m2, m3)
```

## Predictions

Example red wine:

Alcohol: 13, Volatile Acidity: 0.32, Sulphates: 0.85, Quality: 7

```{r echo=F, message=F, warning=F}
thisWine = data.frame(alcohol = 13, volatile.acidity = 0.32,
                       sulphates = 0.85)

predict(m3, newdata = thisWine, interval="prediction", level = .95)
```

It appears that a model that includes alcohol, volatile acidity, and sulphates 
can be used to predict the quality of a wine. The wine that was used to test 
this model had a quality rating of 7. The model predicted a rating of 6.816987. 
One limitation of this model is that since it uses linear regression, it is 
sensitive to outliers. For example, there may be cases were a good wine may have 
a relatively low alcohol conent, which would affect the accuracy of this model. 

------

# Final Plots and Summary

### Volatile Acidity and Wine Quality

```{r echo=F, message=F, warning=F, Plot_One}
ggplot(aes(x = volatile.acidity, y = (..count../sum(..count..) * 100)),
       data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks  = seq(0, 1.6, .1)) + 
  scale_y_continuous(breaks  = seq(0, 7, 1)) + 
  coord_cartesian(xlim = c(0, 1.6)) +  
  ggtitle('Volatile Acidity by Wine Quality') +
  xlab("Volatile Acidity (g / dm^3)") +
  ylab("% of wines with given volatile acidity") + labs(color="Quality") +
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), 
                     panel.border = element_blank(), 
                     panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), 
axis.line = element_line(colour = "black")) + 
  scale_colour_manual(values = c("#A60000", "#0047A6"))
```

Most wines with a volatile acidity less than 0.6 tend to be of good 
quality. Wines with a volatile acidity higher than 0.6 were more likely to be of 
bad quality.

### Alcohol Content and Wine Quality

```{r echo=F, message=F, warning=F, Plot_Two}
ggplot(aes(x = alcohol, y = (..count../sum(..count..) * 100)), data = wine) + geom_freqpoly(aes(color = binary.quality)) + 
  scale_x_continuous(breaks =seq(8,15, 0.5))  + 
  scale_y_continuous(breaks  = seq(0, 12, 1)) +  
  ggtitle('Alcohol content by Wine Quality') + xlab("Alcohol (% by volume)") +
  ylab("% of wines with given alcohol content") + labs(color="Quality") + 
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), 
                     panel.border = element_blank(), 
                     panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), 
axis.line = element_line(colour = "black")) + 
  scale_colour_manual(values = c("#A60000", "#0047A6"))
```

Most wines with an alcohol content greater than 10% are of good quality. There 
are significantly more wines with alcohol contents between 9 and 10% that are of 
bad quality. 

### Fixed Acidity, Alcohol Content, and Wine Quality

```{r echo=F, message=F, warning=F, Plot_Three}
ggplot(data = wine, aes(x = volatile.acidity, y = alcohol)) + 
  geom_point(alpha = 1/3, aes(color = binary.quality)) + 
  scale_x_continuous(breaks =seq(0, 1.5, .2))  + 
  scale_y_continuous(breaks  = seq(0, 16, 2)) +   
  ggtitle('Alcohol Content vs. Volatile Acidity by Wine Quality') +
  xlab("Volatile Acidity (g / dm^3)") +
  ylab("Alcohol (% by volume)") + labs(color="Quality") + theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), 
axis.line = element_line(colour = "black")) + 
  scale_colour_manual(values = c("#A60000", "#0047A6"))
```

Wines with higher alcohol contents and lower volatile acidity levels were more 
likely to be of good quality. Most wines with alcohol contents greater than 10% 
and volatile acidity levels lower than 0.6 (g / dm^3) are of good quality. 

------

# Reflection

Exploring the wine dataset revealed a relationship between the volatile acidity, 
alcohol content, and quality of a wine. A univariate analysis of these variables 
revealed skewed distributions, which seem to be a function of the underlying 
quality of the wines. Prior to conducting a bi-variate and multivariate 
analysis, the wines were divide into two groups(good/bad). Encoding the wines 
with a binary variable helped to discern which features had the most influence 
over the quality of wines. The higher the alcohol content of a wine, the more 
likely it is of good quality. Good quality wines also have lower volatile 
acidity levels. A mulivariate analysis of these features suggested that the 
quality of a wine could be predicted based on volatile acidity and alcohol 
levels. A linear regression model was constructed to predict the quality of a 
wine based off its alcohol content, sulphates, and volatile acidity.Further work 
may be done to assess the accuracy of this model. A larger dataset may reveal 
other findings that were not included in this study. 

# Citations

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.